Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2014-07-17 cordillera-cycle.tex
Rework section 4.3 (missing parts).

\subsubsection{Timing of glaciation}

@@

\subsubsection{Ice configuration during MIS~2}

@@

\subsubsection{Ice configuration during MIS~4}

@@

In the North American Cordillera alike other glaciated regions, the large
majority of the glacial geologic record relates to the last few millennia of
glacier cover, most of the older evidence having been overridden by ice retreat
retreat during the deglaciation\footnote{\textbf{Johan,} can we refer to one
of your papers here?}.

\todo{add 1 sentence to connect to the next paragraph: that is easy:D}

The timing of peak ice volume during the last glacial maximum and the pacing of
deglaciation depend cirticially on the choice of atmospheric forcing
(Figs.~\ref{fig:timeseries} and~\ref{fig:deglacseries}). Adopting the EPICA
atmospheric forcing yields peak ice volume at 17.3\,ka and an uninterupted
deglaciation until 10.6\,ka (Fig.~\ref{fig:deglacseries}, lower panel, red
curves). The simulation driven by the GRIP palaeo-temperature record yields
peak ice volume at 19.1\,ka and a deglaciation interupted by two standstills or
readvances until 9.4\,ka. The first interuption occurs between 16.6 and
14.5\,ka, and the second between 12.6 and 11.6\,ka
(Fig.~\ref{fig:deglacseries}, lower panel, blue curve). Hence, the two model
runs, while similar in overall timing compared to runs with other climate
drivers, differ in detail in that the EPICA depicts peak glaciation almost
2\,ka later and shows a faster uninterrupted deglaciation which yields ice-free
conditions more than 1\,ka earlier.

\todo{From the timing of max glaciation and the duration of the deglaciation,
is there any evidence to prefer one model output above the other? If so, state
that, if not, state that too. The behavior of both models is perhaps reasonable
within the uncertainty envelope that the field evidence provides...?}

\subsubsection{Possibility for a late-glacial readvance}

The possibility of late glacial readvances in the North American Cordillera has
been debated for some time \citep{Osborn.Gerloff.1997}, and locally these have
been reconstructed and dated. Radiocarbon-dated end moraines in the Fraser and
Squamish valleys off the southern tip of the Coast Mountains, indicates
consecutive glacier maximal (or standstills while in overall retreat), one of
which one corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
Kovanen.Easterbrook.2002}. In the Finlay River area of northern Rocky
Mountains, sharp-crested moraines indicate a late-glacial readvance of alpine
glaciers and their interaction with the main body of the decaying Cordilleran
ice sheet \citep{Lakeman.etal.2008}. Additional evidence for late-glacial
alpine glacier readvances includes moraines in the eastern Coast Mountains,
Rocky Mountains and the Columbia Mountains \citep{Osborn.Gerloff.1997,
Menounos.etal.2008}. Although further work is needed to constrain the timing of
the late-glacial readvance, to assess its extents and geographical
distribution, and to identify the potential climatic triggers
\citep{Menounos.etal.2008}, it is interesting to note that the simulation
driven by the GRIP record produces a late-glacial readvance in the Coast
Mountains, Rocky Mountains and the Finlay River area, corresponding to where it
has been identified in the geological record (Fig.~\ref{fig:deglac}, left
panel). In contrast the EPICA-driven simulation produces a nearly-continuous
deglaciation with only a highly restricted glacial readvance
(Fig.~\ref{fig:deglac}, right panel).

\todo{Note that most of the evidence for re-advance is for small glaciers,
      whereas the model output indicates that there was still a big body of ice
      at that time, as in \citet{Lakeman.etal.2008}.}

Predicted patterns of ice sheet retreat are relatively consistent between the
two simulations (Fig.~\ref{fig:deglac}). The southern sector of the modelling
domain, including the Puget Lowland, the Coast and the Rocky mountains, and the
Interior Plateau of British Columbia, becomes completely deglaciated by 12\,ka,
whereas a significant ice cover remains over the Skeena, Selwyn, MacKenzie,
Wrangell and Saint-Elias mountains in the northern sector of the modelling
domain. After 12\,ka, deglaciation continues to proceed across the Liard
Lowland with a radial ice margin retreat towards the surrounding mountain
ranges, and inward flow towards the depression, consistent with the regional
melt water record of the last deglaciation \citep{Margold.etal.2013}
\todo{rephrase with ``unzip''?}. Remaining ice continues to decay by retreating
towards the Selwyn and Skeena mountains. The last remnants of the Cordilleran
ice sheet finally disappears from the Skeena mountains around 10.6\,ka (EPICA)
and 9.4\,ka (GRIP).

@@

% fig:timeseries
\begin{figure*}
  \includegraphics{timeseries}
  \caption{Temperature offset time-series from ice core and sediment core
           records (Table~\ref{tab:records}) used as palaeo-climate forcing for
           the ice sheet model \textbf{(top)}, and modelled ice volume
           through the last 120\,kyr, expressed in meters of sea-level
           equivalent \textbf{(bottom)}. Gray spans indicate Marine Isotope
           Stages (MIS) according to a global compilation of benthic
           \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}. Hatched
           rectangles highlight modelled ice volume extrema corresponding to
           MIS~4 (61.9--55.4\,kyr), MIS~3 (52.2--45.6\,kyr), and
           MIS~2 (last glacial maximum, 29.5--16.9\,kyr). Dotted lines
           correspond to the GRIP and EPICA 6\,km-resolution runs.}
  \label{fig:timeseries}
\end{figure*}

@@

% fig:deglacseries
\begin{figure}
  \includegraphics{deglacseries}
  \caption{Temperature offset time-series from the GRIP and EPICA ice core
           records (Table~\ref{tab:records}) \textbf{(top)}, and modelled ice
           volume during the deglaciation, expressed in meters of sea-level
           equivalent \textbf{(bottom)}.}
  \label{fig:deglacseries}
\end{figure}

% fig:deglacshots
\begin{figure*}
  \includegraphics{deglacshots}
  \caption{Snapshots of modelled surface topography (200\,m contours)
           and surface velocity (colour mapping) during the last deglaciation
           from the GRIP (top panels) and EPICA (bottom panels) simulations.}
  \label{fig:deglacshots-grip}
\end{figure*}

>>>>

\subsubsection{Timing of glaciation}
\label{sec:timing}

@@

\subsubsection{Ice configuration during MIS~2}
\label{sec:mis2}

@@

\subsubsection{Ice configuration during MIS~4}
\label{sec:mis4}

@@

In the North American Cordillera alike other glaciated regions, the large
majority of the glacial geologic record relates to the last few millennia of
glaciation, most of the older evidence having been overridden by ice retreat
during the deglaciation \citep{Kleman.1994, Kleman.etal.2010}. From a
numerical modelling perspective, phases of glacier retreat are more challenging
than phases of growth, because they involve more rapid fluctuations of the ice
margin, increased flow velocities and longitudinal stress gradients, and poorly
understood hydrological processes, typically included in the models through
simple parametrisations \citep[e.g.][]{Clason.etal.2012, Clason.etal.2014,
Bueler.Pelt.2014}, if included at all. However, after mapped moraines
indicative of maximal extents (Sect.~\ref{sec:mis2} and~\ref{sec:mis4}),
geomorphologically-reconstructed patterns of the last deglaciation provide the
second best source of evidence for validation of our simulations.

In the North American Cordillera, the presence of lateral melt-water channels
at high elevation \citep{Margold.etal.2011, Margold.etal.2013a,
Margold.etal.2014}, and abundant esker systems at low elevation
\citep{Burke.etal.2012, Burke.etal.2012a, Perkins.etal.2013,
Margold.etal.2013}\footnote{
    Is this a good choice of references?}
indicate a rapid deglaciation. The southern and northern margins of the
Cordilleran ice sheet reached their last glacial maximum extent around
17\,ka \citep[Sect.~\ref{sec:timing};][]{Porter.Swanson.1998, Cosma.etal.2008,
Stroeven.etal.2010, Stroeven.etal.2014}, putting an older bound to the
onset of ice retreat. However, he timing of the late deglaciation is less well
constrained, but recent cosmogenic dates from north-central British Columbia
indicate that a seizable ice cap emanating from the central Coast Mountains or
the Skeena Mountains persisted into the Younger Dryas chronozone, at least
until 12.4\,ka \citep{Margold.etal.2014}.

In our simulations, the timing of peak ice volume during the last glacial
maximum and the pacing of deglaciation depend cirticially on the choice of
atmospheric forcing (Table~\ref{tab:extrema}, Figs.~\ref{fig:timeseries}
and~\ref{fig:deglacseries}). Adopting the EPICA
atmospheric forcing yields peak ice volume at 17.3\,ka and an uninterupted
deglaciation until 10.6\,ka (Fig.~\ref{fig:deglacseries}, lower panel, red
curves). The simulation driven by the GRIP palaeo-temperature record yields
peak ice volume at 19.1\,ka and a deglaciation interrupted by two ice volume
standstills until 9.4\,ka. The first interuption occurs between 16.6 and
14.5\,ka, and the second between 12.6 and 11.6\,ka
(Fig.~\ref{fig:deglacseries}, lower panel, blue curve). Hence, the two model
runs, while similar in overall timing compared to runs with other climate
drivers, differ in detail in that the EPICA depicts peak glaciation almost
2\,ka later than the GRIP, in closer agreement to dated maximum extents, and
shows a faster, uninterrupted deglaciation which yields ice-free conditions
more than 1\,ka earlier.

Modelled patterns of ice sheet retreat are relatively consistent between the
two simulations (Figs.~\ref{fig:deglacshots} and.~\ref{fig:deglac}). The
southern sector of the modelling domain, including the Puget Lowland, the Coast
and Rocky mountains, and the Interior Plateau of British Columbia, becomes
completely deglaciated by 12\,ka, whereas a significant ice cover remains over
the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias mountains in the
northern sector of the modelling domain. After 12\,ka, deglaciation continues
to proceed across the Liard Lowland with a radial ice margin retreat towards
the surrounding mountain ranges, consistent with the regional melt water record
of the last deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
decay by retreating towards the Selwyn and Skeena mountains. The last remnants
of the Cordilleran ice sheet finally disappear from the Skeena mountains around
10.6\,ka (EPICA) and 9.4\,ka (GRIP).

\subsubsection{Late-glacial readvance}

The possibility of late glacial readvances in the North American Cordillera has
been debated for some time \citep{Osborn.Gerloff.1997}, and locally these have
been reconstructed and dated. Radiocarbon-dated end moraines in the Fraser and
Squamish valleys, off the southern tip of the Coast Mountains, indicate
consecutive glacier maxima, or standstills while in overall retreat, one of
which corresponding to the Younger Dryas chronozone \citep{Clague.etal.1997,
Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
Kovanen.Easterbrook.2002}. Although these moraines characterise independent
valley glaciers, not necessarily connected to an ice sheet, the Finlay River
area of the northern Rocky Mountains presents a different kind of evidence.
There, sharp-crested moraines indicate a late-glacial readvance of local alpine
glaciers and, more importantly, their interaction with larger, lingering
remnants of the main body of the Cordilleran ice sheet in the valleys
\citep{Lakeman.etal.2008}. Additional evidence for late-glacial
alpine glacier readvances includes moraines in the eastern Coast Mountains,
Rocky Mountains and the Columbia Mountains \citep{Osborn.Gerloff.1997,
Menounos.etal.2008}.

Although further work is needed to constrain the timing of the late-glacial
readvance, to assess its extents and geographical distribution, and to identify
the potential climatic triggers \citep{Menounos.etal.2008}, it is interesting
to note that the simulation driven by the GRIP record produces a late-glacial
readvance in the Coast Mountains, Rocky Mountains and the Finlay River area,
corresponding to where it has been identified in the geological record
(Fig.~\ref{fig:deglac}, left panel). In addition to matching the location of
local readvances, the GRIP-driven simulation shows that larger ice bodies
emanating from the Skeena and MacKenzie mountains may still have exist at the
time of this late-glacial readvance\renote{
    Thanks Arjen for your suggestion to reformulate. Good point.}.
In contrast, the EPICA-driven simulation produces a nearly-continuous
deglaciation with only a tightly restricted glacial readvance
(Fig.~\ref{fig:deglac}, right panel).

@@

% fig:timeseries
\begin{figure*}
  \includegraphics{timeseries}
  \caption{Temperature offset time-series from ice core and sediment core
           records (Table~\ref{tab:records}) used as palaeo-climate forcing for
           the ice sheet model (top panel), and modelled ice volume
           through the last 120\,kyr, expressed in meters of sea-level
           equivalent (bottom panel). Gray spans indicate Marine Isotope
           Stages (MIS) according to a global compilation of benthic
           \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}. Hatched
           rectangles highlight modelled ice volume extrema corresponding to
           MIS~4 (61.9--55.4\,kyr), MIS~3 (52.2--45.6\,kyr), and
           MIS~2 (last glacial maximum, 29.5--16.9\,kyr). Dotted lines
           correspond to the GRIP and EPICA 6\,km-resolution runs.}
  \label{fig:timeseries}
\end{figure*}

@@

% fig:deglacseries
\begin{figure}
  \includegraphics{deglacseries}
  \caption{Temperature offset time-series from the GRIP and EPICA ice core
           records (Table~\ref{tab:records}) (top panel), and modelled ice
           volume during the deglaciation, expressed in meters of sea-level
           equivalent (bottom panel).}
  \label{fig:deglacseries}
\end{figure}

% fig:deglacshots
\begin{figure*}
  \includegraphics{deglacshots}
  \caption{Snapshots of modelled surface topography (200\,m contours)
           and surface velocity (colour mapping) during the last deglaciation
           from the GRIP (top panels) and EPICA (bottom panels) simulations.}
  \label{fig:deglacshots}
\end{figure*}
